---
title: "PersonalMovement"
author: "D. Cucchiara"
date: "`r Sys.Date()`"
output: html_document
---
# **Personal Movement Project**
## Reproducible Research Project 1

Data Loading and Pre-Processing:
``` {r reading, cache=TRUE}
rawData <- read.csv("activity.csv", header=TRUE)
# examine the data set
head(rawData)
str(rawData)
dim(rawData)
summary(rawData)
head(rawData)
```
Here, we could transform the date to POSIXct class if needed but it seems
unnecessary.

## What is the mean total number of steps taken per day?
Calculate the total number of steps taken per day:
``` {r subset}
steps <- aggregate(steps ~ date, data = rawData, sum)
# Create histogram
hist(steps$steps, main="Total No. of Steps Taken Per Day", xlab="Total No. Steps",
     ylab="Frequency", col=c("blue2", "cyan1", "darkorchid1", "limegreen", "gold1"))
```
Calculate the mean and median:
``` {r mean}
# mean
mean(steps$steps, na.rm=TRUE)
#median
median(steps$steps, na.rm=TRUE)
```
The mean number of steps taken per day is 10766.19. 
The median number of steps taken per day is 10765.
``` {r}
head(steps)
```

## What is the average daily activity pattern?
Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
``` {r}
# subset the data
stepsIntrv <- aggregate(steps~interval, data = rawData, mean, na.rm = TRUE)
#plot the time series
plot(stepsIntrv$interval, stepsIntrv$steps, type = "l", 
     main = "Average No. Steps per Interval", xlab = "5 minute Interval", 
     ylab = "Average No. Steps Taken")
```
Which 5-minute interval, on average across all the days in the dataset, 
contains the maximum number of steps?
``` {r}
stepsIntrv[which.max(stepsIntrv[,2]),1]
head(stepsIntrv)
```
The maximum average steps occurred in interval 835.

## Imputing missing values
```{r}
# Calculate and report the total number of missing values in the dataset 
numNas <- sum(is.na(rawData$steps))
# Find the mean 
intvMean <- mean(stepsIntrv$steps)
numNas # get total NAs
intvMean # get mean steps/interval

```
There are 2304 missing values in the data set.

Impute the missing values
```{r}
# Create a new dataset that is equal to the original dataset but 
# with the missing data filled in
newSum <- rawData
newSum$steps[is.na(newSum$steps)] <- mean(rawData$steps, na.rm=TRUE)

head(newSum)
```
Create histogram with new data and get mean and median
```{r}
# Make a histogram of the total number of steps taken each day
newSum1 <- aggregate(newSum$steps ~ newSum$date, FUN=sum)
colnames(newSum1) <- c("Date", "Steps")
hist(newSum1$Steps, main="Total No. of Steps Taken Per Day", 
     sub="Missing Values imputed using mean", 
     xlab="Total No. Steps", ylab="Frequency", 
     col=c("blue2", "cyan1", "darkorchid1", "limegreen", "gold1"))
# mean
mean(newSum1$Steps, na.rm=TRUE)
#median
median(newSum1$Steps, na.rm=TRUE)
# Examine new data
head(newSum1)
```
Do these values (mean and median) differ from the estimates from the first 
part of the assignment? *The mean is the same, The median in the raw data*
*was 10765 and now it is also 10766.19.* *The median changed slightly.*
What is the impact of imputing missing data on the estimates of the total 
daily number of steps? *very little impact.*

## Are there differences in activity patterns between weekdays and weekends?
Create a new factor variable in the dataset with two levels – 
“weekday” and “weekend” indicating whether a given date is a weekday or 
weekend day.
```{r}

newSum$DayType <- weekdays(as.Date(newSum$date))

newSum$DayType <- ifelse(newSum$DayType %in% c('Saturday','Sunday'),'Weekend','Weekday')

head(newSum)
```
```{r}
# Open necessary  packages
library(lattice) 
library(plyr)
# Get data make graph
newSumInt <- ddply(newSum, .(interval, DayType), summarize, avg = mean(steps))
xyplot(avg~interval|DayType, data=newSumInt, type="l", layout=c(1,2),
       main="Average Steps per Interval by Day Type (Weekend or Weekday)", 
       xlab="Interval", ylab = "Average No. Steps", col="red")
```







